The effect of climate forcing on numerical simulations of the Cordilleran Ice Sheet at the Last Glacial Maximum

### 2014-03-19 cordillera-climate.tex
Typesetting: unbreak post-article space

% Abstract
\begin{abstract}
We present an ensemble of numerical simulations of the Cordilleran Ice Sheet during the Last Glacial Maximum performed with the Parallel Ice Sheet Model (PISM), applying temperature offsets to the present-day climatologies from five different datasets. Monthly mean surface air temperature and precipitation from WorldClim, the NCEP/NCAR reanalysis, the ERA-Interim reanalysis, the Climate Forecast System Reanalysis and the North American Regional Reanalysis are used to compute surface mass balance in a positive degree-day model. Modelled ice sheet outlines and volumes appear highly sensitive to the choice of climate forcing. For three of the four reanalysis datasets used, differences in precipitation are the major source for discrepancies between model results. We assess model performance against a geomorphological reconstruction of the ice margin at the Last Glacial Maximum, and suggest that part of the mismatch is due to unresolved orographic precipitation effects caused by the coarse resolution of reanalysis data. The best match between model output and the reconstructed ice margin is obtained using the high-resolution North American Regional Reanalysis, which we retain for simulations of the Cordilleran Ice Sheet in the future.
\end{abstract}

@@

At the Last Glacial Maximum (LGM), glaciers of a size comparable to the present Greenland and Antarctic ice sheets covered parts of Northern America (Laurentide, Cordilleran and Innuitian ice sheets) and northern Eurasia (Fennoscandian Ice Sheet). Numerical modelling of these former ice masses allows for a comparison between glaciological theories embedded in the models and geomorphological traces underpinning palaeo-glaciological reconstructions. Yet, a major obstacle in this exercise resides in large uncertainties concerning the climate forcing, typically atmospheric temperature and precipitation, required as input to numerical glacier models \citep{hebeler-etal-2008}. This includes uncertainty in the representation of Earth's present climate in regions of poor station coverage, and even larger uncertainty concerning accurate reconstructions of past climate change.

Arguably the most physically sound way to force an ice sheet model for simulations of glacial history is to couple it with a General Circulation Model \citep[GCM;][]{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013}. However the computational demand of GCMs is such that only models of intermediate complexity can run on the time-scales of tens of thousands of years characteristic of ice sheet growth and decay.

Climatologies obtained from GCM palaeo-climate simulations such as produced within the Paleoclimate Modelling Intercomparison Project \citep[PMIP;][]{joussaume-taylor-1995} provide perhaps a more reasonable representation of past climate. However, as such climatologies are only available for specific periods of time, using them as climate forcing for an ice sheet model requires either an assumption of steady-state ice sheet response to climatic fluctuations \citep{huybrechts-tsiobbel-1996}, or interpolation through time between climatologies from different periods, which can be linear \citep{charbit-etal-2002}, or modulated by a ``glacial index'' weighting function derived from ice core $\delta^{18}$O records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}. An important drawback in this approach is that GCM palaeo-climate simulations themselves rely on an accurate description of the global surface topography and therefore include global ice sheet reconstructions such as ICE-4G \citep{peltier-1994} for their surface topographic boundary condition. This condition could potentially exert influence on subsequently modelled ice sheet geometries, and particularly glacier extent.

@@

Here, we propose to address some of the uncertainties concerning climate forcing of numerical glacier models by evaluating the responses of a numerical model, in terms of glacier extent, to inputs from several climate datasets. This is not entirely a new approach as a few studies of this kind are presently available. \citet{quiquet-etal-2012}, for example, assessed the sensitivity of a Greenland ice sheet model to various climate forcing fields, including a regional parametrization by \citet{fausto-etal-2009a}, output from several GCMs and an atmospheric reanalysis. \citet{rodgers-etal-2004} and \citet{charbit-etal-2007} tested the sensitivity of a model of Northern Hemisphere ice sheets to climate forcing from different PMIP LGM simulations. Although using different set-ups, all three studies demonstrate the very large sensitivity of ice sheet models to the choice of climate forcing data used. To limit the degrees of freedom in our model and obtain results independent of palaeo-ice sheet reconstructions such as ICE-4G, we use a simple temperature offset similar to the approaches by \citet{greve-etal-1999} and \citet{bintanja-etal-2005} and assess ice sheet model sensitivity to the choice of present-day climate data. Hence, rather than using GCM output, we force our model with climate reanalysis data, which builds on observational information through data assimilation \citep{bengtsson-etal-2007}. Furthermore, we focus our study regionally on the former Cordilleran Ice Sheet in western North America.

The Cordilleran Ice Sheet (Fig.~\ref{fig:locmap}) covered an area that presently experiences strong regional variations in climate. From a numerical modelling perspective, it is one of the least studied palaeo-ice sheets of the Northern Hemisphere, despite the fact that significant stratigraphical, geomorphological, and chronological data is available to constrain its extent \citep{jackson-clague-1991,porter-swanson-1998,dukrodkin-1999,dyke-2004,kaufman-manley-2004,kleman-etal-2010,stroeven-etal-2010,stroeven-etal-inpress,margold-etal-2011}. The Cordilleran Ice Sheet has previously been modelled as part of efforts to simulate ice sheets in North America \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-1997,tarasov-peltier-2004,gregoire-etal-2012}, the Northern Hemisphere \citep{huybrechts-tsiobbel-1996,greve-etal-1999,charbit-etal-2002,charbit-etal-2007,charbit-etal-2013,johnson-fastook-2002,rodgers-etal-2004,bintanja-etal-2005,zweck-huybrechts-2005,abeouchi-etal-2007} and world-wide \citep{yoshimori-etal-2001}. While these studies generally reproduce the magnitude of North American glaciation at LGM reasonably well, there exists a tendency in the simulations that are independent of ice sheet reconstructions such as the ICE-4G to predict excessive ice cover in parts of northern Yukon Territory and interior Alaska that have remained ice-free throughout the Pleistocene \citep{dukrodkin-1999,kaufman-manley-2004}.

Here we use PISM, a Parallel Ice Sheet Model \citep{web:pism}, to simulate the extent and thickness of the Cordilleran Ice Sheet at the LGM (Fig.~\ref{fig:locmap}). We force our model with multiple climate datasets and compare the modelled ice extent to a geomorphological reconstruction of the LGM ice sheet margin by \citet{dyke-2004}. We aim to determine the climate dataset which is most suited for simulation of the Cordilleran Ice Sheet and to use as input for future, transient studies over a glacial cycle.

@@

As we aim to model glacial inception and growth of the Cordilleran Ice Sheet towards a configuration as last attained during the LGM, we mimic palaeo-climatic conditions by applying constant temperature offsets homogeneously over the modelling domain. Each simulation starts from ice-free conditions and runs for 10\,ka, a time interval representative of the rapid build-up of the last Cordilleran Ice Sheet from nearly ice-free to full glacial conditions \citep{clague-1989,stroeven-etal-2010}. Our simulations are performed in parallel on 32 cores at the Swedish National Supercomputing Center.

@@

The central part of an ice sheet model consists of the computation of flow velocity which itself depends on temperature. PISM is a shallow model, which implies that the balance of stresses is approximated based on their predominant components. On the other hand, the model is polythermal: it accounts for differences in temperature and softness within the ice column.

The Shallow Shelf Approximation (SSA) is used as a ``sliding law'' for the Shallow Ice Approximation (SIA) \citep{bueler-brown-2009,winkelmann-etal-2011}. SIA and SSA velocities are computed by finite difference methods on a 10\,km-resolution horizontal grid of 300 by 150 points (the modelling domain). Ice softness depends on temperature and water content through an enthalpy formulation \citep{aschwanden-blatter-2009,aschwanden-etal-2012}. Enthalpy is computed in three dimensions in up to 51 irregularly spaced layers within the ice, and temperature is additionally computed in 11 regularly spaced layers in bedrock to a depth of 1\,km. Surface air temperature from the climate forcing provides the upper boundary condition to the ice enthalpy model, and a geothermal heat flux of 70\,\unit{mW\,m^{-2}} provides the lower boundary condition to the bedrock thermal model. Although this uniform value does not account for the high spatial geothermal variability in the region, it is on average representative of available heat flow measurements \citep{artemieva-mooney-2001,blackwell-richards-2004}. Whereas spatial variations in the geothermal heat flux may significantly affect basal conditions \citep{pattyn-2010}, this study focuses on reconstructing the former extent of the Cordilleran Ice Sheet, on which it has relatively little effect \citep{rogozhina-etal-2012}.

A pseudo-plastic sliding law \citep{aschwanden-etal-2013} relates the bed-parallel shear stress and the sliding velocity. The yield stress is modelled using the Mohr-Coulomb criterion. The till friction angle $\phi$ varies from 10 to 30\degree\ and is a function of current bed elevation, with lowest values occurring at low elevations,

@@

where $\phi$ is given in degrees and $z$ in meters above current sea level. This model thus accounts for a weakening of till associated with the presence of marine sediments \citep{martin-etal-2011,aschwanden-etal-2013}. Basal topography (Fig.~\ref{fig:topo}) is derived from the ETOPO1 combined topography and bathymetry dataset with a resolution of 1\,arc-minute \citep{data:etopo1}. Sea level is lowered by 120\,m and basal topography responds to ice load following a regional isostasy model that includes lithospheric flexure and mantle relaxation \citep{lingle-clark-1985}.

@@

Ice surface accumulation and ablation are computed from monthly mean surface air temperature and monthly precipitation by a temperature-index (positive degree-day) model \citep{hock-2003}. Ice accumulation is equal to precipitation when temperature is below 0\,\unit{\degree C}, and decreases to zero linearly with temperature between 0 and 2\,\unit{\degree C}. Ice ablation is computed from the number of positive degree-days, defined as the integral of temperatures above 0\,\unit{\degree C} in one year.

The positive degree-day integral \citep{calov-greve-2005} is numerically approximated using week-long sub-intervals. It accounts for temperature variability assuming a normal distribution along a central (input) value. The temperature standard deviation is a constant model parameter and was assigned a value of 3.07\,\unit{\degree C}, which corresponds to the mean summer (JJA), model domain-averaged monthly standard deviation of daily mean temperature from monthly mean temperature, as computed from North American Regional Reanalysis data \citep{data:narr} in a manner similar to \citet{seguinot-2013}. The ablation model incorporates degree-day factors of 3.04\,\unit{mm\,K^{-1}\,day^{-1}} for snow and 4.59\,\unit{mm\,K^{-1}\,day^{-1}} for ice, as derived from mass-balance measurements on contemporary glaciers from the Coast Mountains and Rocky Mountains in British Columbia \citep{shea-etal-2009}.

@@

Prior to surface mass balance computation, the model dynamically applies a lapse-rate correction to surface air temperature. This correction accounts for the evolution of ice thickness on the one hand, and differences between the climate forcing reference topography and the ice flow model basal topography on the other hand. It uses a reference topography distinct from the modelled basal topography and specific to each climate forcing dataset (Fig.~\ref{fig:topo}). An annual lapse rate of 6\,\unit{\degree C\,km^{-1}} is used in all simulations. No corrections apply to precipitation changes with elevation.

@@

WorldClim is a high resolution global climate dataset built from meteorological observations \citep{data:worldclim}. It was built by applying a spatial interpolation between a large set of measurements taken at terrestrial weather stations worldwide, using SRTM \citep{data:srtm} and GTOPO30 elevation data \citep{data:gtopo30}. This dataset provides a resolution much higher than attained by circulation models, however its reliability is variable due to the highly inhomogeneous distribution of measurements \citep{data:worldclim}.

Within our modelling domain, the spatial distributions of summer (JJA) and winter (DJF) air surface temperatures in WorldClim generally show a decrease of temperature from south to north (JJA) and from south-west to north-east (DJF), such that regions further inland experience colder winters (Fig.~\ref{fig:temp}). It should be noted that temperature gradients are much stronger in winter than in summer. Temperature rises well above zero during the summer months over the entire modelling domain, except for the highest mountain peaks, and even regions where mean annual temperatures are well below freezing point do experience warm summers. In other words, there is a strong contrast in seasonality between coastal temperate-climate and inland continental-climate regions.

The spatial distribution of JJA and DJF mean precipitation rates in WorldClim reveals a strong precipitation decline beyond the coastal regions (Fig.~\ref{fig:prec}), primarily as a result of the continuous orographic barrier formed by the Wrangell and St.-Elias mountains, the Coast Mountains and the Cascades (Fig.~\ref{fig:locmap}). A less pronounced precipitation peak is also visible over the interior ranges. From an ice sheet mass balance point of view, these heterogeneities become even stronger due to the difference in timing of the precipitation peak through the year. Whereas coastal regions experience most precipitation as snow during the accumulation season (DJF), inland regions experience dry winters and most of the annual precipitation consequently falls as rain during the ablation season (JJA).

@@

The CFSR climatology presents wave-like artefacts in its precipitation field (Fig.~\ref{fig:prec}). These are a common feature in GCM simulations due to the failure of spectral models to resolve the very local nature of topography-induced precipitation events, which becomes particularly visible in the high-resolution, less spatially-smoothed data. Therefore, an alternative forcing was prepared for CFSR by smoothing its precipitation field. This was achieved by averaging data locally in a circular kernel of 7\,pixels in diameter prior to re-projection using the module r.neighbors from GRASS~GIS \citep{soft:grass}. Finally, input files for PISM were prepared using \citet{web:nc4py}.

@@

Across the range of temperature offsets used, final glaciated area and final ice volume differ widely between different input climatologies (Fig.~\ref{fig:ivolarea}). For instance, using a 5\,\unit{\degree C} temperature offset, final glaciated area differs between WorldClim and CFSR forcing by a factor of~6 and final volume by a factor of~12 (Fig.~\ref{fig:ivolarea}). For particular temperature offset values, given input climatologies result in similar final glaciated areas and ice volumes, yet regional patterns are different (Fig.~\ref{fig:extent}).

Under present climate (0\,\unit{\degree C} offset), WorldClim, ERA-Interim and NARR forcing produce local ice caps in areas presently glaciated (Fig.~\ref{fig:extent}). These results are fairly consistent with contemporary ice distribution, although both the 10~km horizontal resolution and the occurrence of glaciers and ice caps surface topographies in the ETOPO1 dataset used for basal topography prevent a meaningful detailed comparison. CFSR forcing, on the other hand, yields continuous ice cover over the northern Coast Mountains, and NCEP/NCAR forcing results in ice sheet growth, both of which are largely inconsistent with the present state of glaciation. At the opposite side of the temperature offset range, it can be noted that both the NCEP/NCAR and CFSR forcing produce oversized ice sheets whose extents are primarily restricted by the domain boundaries (Fig.~\ref{fig:extent}). The smoothing of CFSR precipitation yields an insignificant effect on the resulting ice sheet geometry.

To contrast the effect of the six different input climatologies, we plot simulations using a 5\,\unit{\degree C} temperature offset (Fig.~\ref{fig:cool05}). The magnitude of glaciation differs dramatically from one climate forcing to another. Whereas the NCEP/NCAR and CFSR climatologies produce a large ice sheet that covers most of the model domain (Fig.~\ref{fig:cool05}, lower panels), WorldClim, ERA-Interim and NARR climatologies lead to a more restrictive ice cover (Fig.~\ref{fig:cool05}, upper panels). The WorldClim run only resulted in the growth of mountain ice caps bounded to the highest topography.

@@

These simulations of the LGM Cordilleran Ice Sheet show a very large sensitivity to the choice of climate forcing data used (Figs.~\ref{fig:ivolarea}--\ref{fig:cool05}). To understand the origin of the discrepancies between model results, we quantify differences in surface air temperature and precipitation rate between the input climatologies and test their effect on the ice sheet model. We then compare our findings to a geomorphological reconstruction of the LGM Cordilleran Ice Sheet margin, and discuss some of the key assumptions of this study.

@@

As previously outlined, significant differences in the distribution of winter (DJF) precipitation and summer (JJA) temperature, which are the predominant controls on accumulation and melt at the modelled ice sheet surface, exist between datasets (Figs.~\ref{fig:topo}--\ref{fig:prec}). In this section we use the WorldClim climatology, which is derived from observations, as a reference with which to compare other climate forcing data on land. ERA-Interim, CFSR and NARR climatologies exhibit temperature distributions fairly consistent with WorldClim data (Fig.~\ref{fig:tempheatmap}). When examining the spatial distribution of temperature differences between reanalyses and WorldClim data, it appears that most of the disparity between reanalyses and observation data is caused by topographical detail at scales unresolved in the reanalysis datasets (Fig.~\ref{fig:tempdiff}). Surface air temperature data from the NCEP/NCAR climatology shows the largest disparity with that of WorldClim, with a significant cold anomaly over most of the modelling domain (Figs.~\ref{fig:tempheatmap}--\ref{fig:tempdiff}).

@@

Most likely, this is the signature of an orographic precipitation effect. As described in section~\ref{sec:climate}, the topography of the Northern American Cordillera is such that its western ranges form a continuous orographic barrier, causing high precipitation along the Pacific coast while leaving much of the interior arid. However the ability of a GCM to reproduce these contrasts in precipitation is bounded by spatial resolution. In a model of coarser resolution, these high mountain ranges are reduced to smooth hills and the modelled distribution of precipitation is also smoother (Fig.~\ref{fig:oroprecip}). However, this effect alone cannot explain the widespread and pronounced positive precipitation anomalies observed in CFSR data (Figs.~\ref{fig:precheatmap}--\ref{fig:precdiff}) despite of its high spatial resolution (Table~\ref{tab:reanalyses}).

@@

To distinguish the effects of temperature and precipitation biases on the ice sheet model, we run a series of eight additional simulations. These new simulations use a “hybrid” climate forcing that consists of temperature data from WorldClim, combined with precipitation data from each of the four reanalyses, and vice-versa. A single temperature offset of 5\,\unit{\degree C} is used. This value was chosen to allow for comparison with previous results (Fig.~\ref{fig:cool05}). In each case, temperature lapse rates are computed using the reference topography from the corresponding dataset.

>>>>

% Abstract
\begin{abstract}
We present an ensemble of numerical simulations of the Cordilleran Ice Sheet during the Last Glacial Maximum performed with the Parallel Ice Sheet Model (PISM), applying temperature offsets to the present-day climatologies from five different datasets. Monthly mean surface air temperature and precipitation from WorldClim, the NCEP/NCAR reanalysis, the ERA-Interim reanalysis, the Climate Forecast System Reanalysis and the North American Regional Reanalysis are used to compute surface mass balance in a~positive degree-day model. Modelled ice sheet outlines and volumes appear highly sensitive to the choice of climate forcing. For three of the four reanalysis datasets used, differences in precipitation are the major source for discrepancies between model results. We assess model performance against a~geomorphological reconstruction of the ice margin at the Last Glacial Maximum, and suggest that part of the mismatch is due to unresolved orographic precipitation effects caused by the coarse resolution of reanalysis data. The best match between model output and the reconstructed ice margin is obtained using the high-resolution North American Regional Reanalysis, which we retain for simulations of the Cordilleran Ice Sheet in the future.
\end{abstract}

@@

At the Last Glacial Maximum (LGM), glaciers of a~size comparable to the present Greenland and Antarctic ice sheets covered parts of Northern America (Laurentide, Cordilleran and Innuitian ice sheets) and northern Eurasia (Fennoscandian Ice Sheet). Numerical modelling of these former ice masses allows for a~comparison between glaciological theories embedded in the models and geomorphological traces underpinning palaeo-glaciological reconstructions. Yet, a~major obstacle in this exercise resides in large uncertainties concerning the climate forcing, typically atmospheric temperature and precipitation, required as input to numerical glacier models \citep{hebeler-etal-2008}. This includes uncertainty in the representation of Earth's present climate in regions of poor station coverage, and even larger uncertainty concerning accurate reconstructions of past climate change.

Arguably the most physically sound way to force an ice sheet model for simulations of glacial history is to couple it with a~General Circulation Model \citep[GCM;][]{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013}. However the computational demand of GCMs is such that only models of intermediate complexity can run on the time-scales of tens of thousands of years characteristic of ice sheet growth and decay.

Climatologies obtained from GCM palaeo-climate simulations such as produced within the Paleoclimate Modelling Intercomparison Project \citep[PMIP;][]{joussaume-taylor-1995} provide perhaps a~more reasonable representation of past climate. However, as such climatologies are only available for specific periods of time, using them as climate forcing for an ice sheet model requires either an assumption of steady-state ice sheet response to climatic fluctuations \citep{huybrechts-tsiobbel-1996}, or interpolation through time between climatologies from different periods, which can be linear \citep{charbit-etal-2002}, or modulated by a~``glacial index'' weighting function derived from ice core $\delta^{18}$O records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}. An important drawback in this approach is that GCM palaeo-climate simulations themselves rely on an accurate description of the global surface topography and therefore include global ice sheet reconstructions such as ICE-4G \citep{peltier-1994} for their surface topographic boundary condition. This condition could potentially exert influence on subsequently modelled ice sheet geometries, and particularly glacier extent.

@@

Here, we propose to address some of the uncertainties concerning climate forcing of numerical glacier models by evaluating the responses of a~numerical model, in terms of glacier extent, to inputs from several climate datasets. This is not entirely a~new approach as a~few studies of this kind are presently available. \citet{quiquet-etal-2012}, for example, assessed the sensitivity of a~Greenland ice sheet model to various climate forcing fields, including a~regional parametrization by \citet{fausto-etal-2009a}, output from several GCMs and an atmospheric reanalysis. \citet{rodgers-etal-2004} and \citet{charbit-etal-2007} tested the sensitivity of a~model of Northern Hemisphere ice sheets to climate forcing from different PMIP LGM simulations. Although using different set-ups, all three studies demonstrate the very large sensitivity of ice sheet models to the choice of climate forcing data used. To limit the degrees of freedom in our model and obtain results independent of palaeo-ice sheet reconstructions such as ICE-4G, we use a~simple temperature offset similar to the approaches by \citet{greve-etal-1999} and \citet{bintanja-etal-2005} and assess ice sheet model sensitivity to the choice of present-day climate data. Hence, rather than using GCM output, we force our model with climate reanalysis data, which builds on observational information through data assimilation \citep{bengtsson-etal-2007}. Furthermore, we focus our study regionally on the former Cordilleran Ice Sheet in western North America.

The Cordilleran Ice Sheet (Fig.~\ref{fig:locmap}) covered an area that presently experiences strong regional variations in climate. From a~numerical modelling perspective, it is one of the least studied palaeo-ice sheets of the Northern Hemisphere, despite the fact that significant stratigraphical, geomorphological, and chronological data is available to constrain its extent \citep{jackson-clague-1991,porter-swanson-1998,dukrodkin-1999,dyke-2004,kaufman-manley-2004,kleman-etal-2010,stroeven-etal-2010,stroeven-etal-inpress,margold-etal-2011}. The Cordilleran Ice Sheet has previously been modelled as part of efforts to simulate ice sheets in North America \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-1997,tarasov-peltier-2004,gregoire-etal-2012}, the Northern Hemisphere \citep{huybrechts-tsiobbel-1996,greve-etal-1999,charbit-etal-2002,charbit-etal-2007,charbit-etal-2013,johnson-fastook-2002,rodgers-etal-2004,bintanja-etal-2005,zweck-huybrechts-2005,abeouchi-etal-2007} and world-wide \citep{yoshimori-etal-2001}. While these studies generally reproduce the magnitude of North American glaciation at LGM reasonably well, there exists a~tendency in the simulations that are independent of ice sheet reconstructions such as the ICE-4G to predict excessive ice cover in parts of northern Yukon Territory and interior Alaska that have remained ice-free throughout the Pleistocene \citep{dukrodkin-1999,kaufman-manley-2004}.

Here we use PISM, a~Parallel Ice Sheet Model \citep{web:pism}, to simulate the extent and thickness of the Cordilleran Ice Sheet at the LGM (Fig.~\ref{fig:locmap}). We force our model with multiple climate datasets and compare the modelled ice extent to a~geomorphological reconstruction of the LGM ice sheet margin by \citet{dyke-2004}. We aim to determine the climate dataset which is most suited for simulation of the Cordilleran Ice Sheet and to use as input for future, transient studies over a~glacial cycle.

@@

As we aim to model glacial inception and growth of the Cordilleran Ice Sheet towards a~configuration as last attained during the LGM, we mimic palaeo-climatic conditions by applying constant temperature offsets homogeneously over the modelling domain. Each simulation starts from ice-free conditions and runs for 10\,ka, a~time interval representative of the rapid build-up of the last Cordilleran Ice Sheet from nearly ice-free to full glacial conditions \citep{clague-1989,stroeven-etal-2010}. Our simulations are performed in parallel on 32 cores at the Swedish National Supercomputing Center.

@@

The central part of an ice sheet model consists of the computation of flow velocity which itself depends on temperature. PISM is a~shallow model, which implies that the balance of stresses is approximated based on their predominant components. On the other hand, the model is polythermal: it accounts for differences in temperature and softness within the ice column.

The Shallow Shelf Approximation (SSA) is used as a~``sliding law'' for the Shallow Ice Approximation (SIA) \citep{bueler-brown-2009,winkelmann-etal-2011}. SIA and SSA velocities are computed by finite difference methods on a~10\,km-resolution horizontal grid of 300 by 150 points (the modelling domain). Ice softness depends on temperature and water content through an enthalpy formulation \citep{aschwanden-blatter-2009,aschwanden-etal-2012}. Enthalpy is computed in three dimensions in up to 51 irregularly spaced layers within the ice, and temperature is additionally computed in 11 regularly spaced layers in bedrock to a~depth of 1\,km. Surface air temperature from the climate forcing provides the upper boundary condition to the ice enthalpy model, and a~geothermal heat flux of 70\,\unit{mW\,m^{-2}} provides the lower boundary condition to the bedrock thermal model. Although this uniform value does not account for the high spatial geothermal variability in the region, it is on average representative of available heat flow measurements \citep{artemieva-mooney-2001,blackwell-richards-2004}. Whereas spatial variations in the geothermal heat flux may significantly affect basal conditions \citep{pattyn-2010}, this study focuses on reconstructing the former extent of the Cordilleran Ice Sheet, on which it has relatively little effect \citep{rogozhina-etal-2012}.

A~pseudo-plastic sliding law \citep{aschwanden-etal-2013} relates the bed-parallel shear stress and the sliding velocity. The yield stress is modelled using the Mohr-Coulomb criterion. The till friction angle $\phi$ varies from 10 to 30\degree\ and is a~function of current bed elevation, with lowest values occurring at low elevations,

@@

where $\phi$ is given in degrees and $z$ in meters above current sea level. This model thus accounts for a~weakening of till associated with the presence of marine sediments \citep{martin-etal-2011,aschwanden-etal-2013}. Basal topography (Fig.~\ref{fig:topo}) is derived from the ETOPO1 combined topography and bathymetry dataset with a~resolution of 1\,arc-minute \citep{data:etopo1}. Sea level is lowered by 120\,m and basal topography responds to ice load following a~regional isostasy model that includes lithospheric flexure and mantle relaxation \citep{lingle-clark-1985}.

@@

Ice surface accumulation and ablation are computed from monthly mean surface air temperature and monthly precipitation by a~temperature-index (positive degree-day) model \citep{hock-2003}. Ice accumulation is equal to precipitation when temperature is below 0\,\unit{\degree C}, and decreases to zero linearly with temperature between 0 and 2\,\unit{\degree C}. Ice ablation is computed from the number of positive degree-days, defined as the integral of temperatures above 0\,\unit{\degree C} in one year.

The positive degree-day integral \citep{calov-greve-2005} is numerically approximated using week-long sub-intervals. It accounts for temperature variability assuming a~normal distribution along a~central (input) value. The temperature standard deviation is a~constant model parameter and was assigned a~value of 3.07\,\unit{\degree C}, which corresponds to the mean summer (JJA), model domain-averaged monthly standard deviation of daily mean temperature from monthly mean temperature, as computed from North American Regional Reanalysis data \citep{data:narr} in a~manner similar to \citet{seguinot-2013}. The ablation model incorporates degree-day factors of 3.04\,\unit{mm\,K^{-1}\,day^{-1}} for snow and 4.59\,\unit{mm\,K^{-1}\,day^{-1}} for ice, as derived from mass-balance measurements on contemporary glaciers from the Coast Mountains and Rocky Mountains in British Columbia \citep{shea-etal-2009}.

@@

Prior to surface mass balance computation, the model dynamically applies a~lapse-rate correction to surface air temperature. This correction accounts for the evolution of ice thickness on the one hand, and differences between the climate forcing reference topography and the ice flow model basal topography on the other hand. It uses a~reference topography distinct from the modelled basal topography and specific to each climate forcing dataset (Fig.~\ref{fig:topo}). An annual lapse rate of 6\,\unit{\degree C\,km^{-1}} is used in all simulations. No corrections apply to precipitation changes with elevation.

@@

WorldClim is a~high resolution global climate dataset built from meteorological observations \citep{data:worldclim}. It was built by applying a~spatial interpolation between a~large set of measurements taken at terrestrial weather stations worldwide, using SRTM \citep{data:srtm} and GTOPO30 elevation data \citep{data:gtopo30}. This dataset provides a~resolution much higher than attained by circulation models, however its reliability is variable due to the highly inhomogeneous distribution of measurements \citep{data:worldclim}.

Within our modelling domain, the spatial distributions of summer (JJA) and winter (DJF) air surface temperatures in WorldClim generally show a~decrease of temperature from south to north (JJA) and from south-west to north-east (DJF), such that regions further inland experience colder winters (Fig.~\ref{fig:temp}). It should be noted that temperature gradients are much stronger in winter than in summer. Temperature rises well above zero during the summer months over the entire modelling domain, except for the highest mountain peaks, and even regions where mean annual temperatures are well below freezing point do experience warm summers. In other words, there is a~strong contrast in seasonality between coastal temperate-climate and inland continental-climate regions.

The spatial distribution of JJA and DJF mean precipitation rates in WorldClim reveals a~strong precipitation decline beyond the coastal regions (Fig.~\ref{fig:prec}), primarily as a~result of the continuous orographic barrier formed by the Wrangell and St.-Elias mountains, the Coast Mountains and the Cascades (Fig.~\ref{fig:locmap}). A~less pronounced precipitation peak is also visible over the interior ranges. From an ice sheet mass balance point of view, these heterogeneities become even stronger due to the difference in timing of the precipitation peak through the year. Whereas coastal regions experience most precipitation as snow during the accumulation season (DJF), inland regions experience dry winters and most of the annual precipitation consequently falls as rain during the ablation season (JJA).

@@

The CFSR climatology presents wave-like artefacts in its precipitation field (Fig.~\ref{fig:prec}). These are a~common feature in GCM simulations due to the failure of spectral models to resolve the very local nature of topography-induced precipitation events, which becomes particularly visible in the high-resolution, less spatially-smoothed data. Therefore, an alternative forcing was prepared for CFSR by smoothing its precipitation field. This was achieved by averaging data locally in a~circular kernel of 7\,pixels in diameter prior to re-projection using the module r.neighbors from GRASS~GIS \citep{soft:grass}. Finally, input files for PISM were prepared using \citet{web:nc4py}.

@@

Across the range of temperature offsets used, final glaciated area and final ice volume differ widely between different input climatologies (Fig.~\ref{fig:ivolarea}). For instance, using a~5\,\unit{\degree C} temperature offset, final glaciated area differs between WorldClim and CFSR forcing by a~factor of~6 and final volume by a~factor of~12 (Fig.~\ref{fig:ivolarea}). For particular temperature offset values, given input climatologies result in similar final glaciated areas and ice volumes, yet regional patterns are different (Fig.~\ref{fig:extent}).

Under present climate (0\,\unit{\degree C} offset), WorldClim, ERA-Interim and NARR forcing produce local ice caps in areas presently glaciated (Fig.~\ref{fig:extent}). These results are fairly consistent with contemporary ice distribution, although both the 10~km horizontal resolution and the occurrence of glaciers and ice caps surface topographies in the ETOPO1 dataset used for basal topography prevent a~meaningful detailed comparison. CFSR forcing, on the other hand, yields continuous ice cover over the northern Coast Mountains, and NCEP/NCAR forcing results in ice sheet growth, both of which are largely inconsistent with the present state of glaciation. At the opposite side of the temperature offset range, it can be noted that both the NCEP/NCAR and CFSR forcing produce oversized ice sheets whose extents are primarily restricted by the domain boundaries (Fig.~\ref{fig:extent}). The smoothing of CFSR precipitation yields an insignificant effect on the resulting ice sheet geometry.

To contrast the effect of the six different input climatologies, we plot simulations using a~5\,\unit{\degree C} temperature offset (Fig.~\ref{fig:cool05}). The magnitude of glaciation differs dramatically from one climate forcing to another. Whereas the NCEP/NCAR and CFSR climatologies produce a~large ice sheet that covers most of the model domain (Fig.~\ref{fig:cool05}, lower panels), WorldClim, ERA-Interim and NARR climatologies lead to a~more restrictive ice cover (Fig.~\ref{fig:cool05}, upper panels). The WorldClim run only resulted in the growth of mountain ice caps bounded to the highest topography.

@@

These simulations of the LGM Cordilleran Ice Sheet show a~very large sensitivity to the choice of climate forcing data used (Figs.~\ref{fig:ivolarea}--\ref{fig:cool05}). To understand the origin of the discrepancies between model results, we quantify differences in surface air temperature and precipitation rate between the input climatologies and test their effect on the ice sheet model. We then compare our findings to a~geomorphological reconstruction of the LGM Cordilleran Ice Sheet margin, and discuss some of the key assumptions of this study.

@@

As previously outlined, significant differences in the distribution of winter (DJF) precipitation and summer (JJA) temperature, which are the predominant controls on accumulation and melt at the modelled ice sheet surface, exist between datasets (Figs.~\ref{fig:topo}--\ref{fig:prec}). In this section we use the WorldClim climatology, which is derived from observations, as a~reference with which to compare other climate forcing data on land. ERA-Interim, CFSR and NARR climatologies exhibit temperature distributions fairly consistent with WorldClim data (Fig.~\ref{fig:tempheatmap}). When examining the spatial distribution of temperature differences between reanalyses and WorldClim data, it appears that most of the disparity between reanalyses and observation data is caused by topographical detail at scales unresolved in the reanalysis datasets (Fig.~\ref{fig:tempdiff}). Surface air temperature data from the NCEP/NCAR climatology shows the largest disparity with that of WorldClim, with a~significant cold anomaly over most of the modelling domain (Figs.~\ref{fig:tempheatmap}--\ref{fig:tempdiff}).

@@

Most likely, this is the signature of an orographic precipitation effect. As described in section~\ref{sec:climate}, the topography of the Northern American Cordillera is such that its western ranges form a~continuous orographic barrier, causing high precipitation along the Pacific coast while leaving much of the interior arid. However the ability of a~GCM to reproduce these contrasts in precipitation is bounded by spatial resolution. In a~model of coarser resolution, these high mountain ranges are reduced to smooth hills and the modelled distribution of precipitation is also smoother (Fig.~\ref{fig:oroprecip}). However, this effect alone cannot explain the widespread and pronounced positive precipitation anomalies observed in CFSR data (Figs.~\ref{fig:precheatmap}--\ref{fig:precdiff}) despite of its high spatial resolution (Table~\ref{tab:reanalyses}).

@@

To distinguish the effects of temperature and precipitation biases on the ice sheet model, we run a~series of eight additional simulations. These new simulations use a~“hybrid” climate forcing that consists of temperature data from WorldClim, combined with precipitation data from each of the four reanalyses, and vice-versa. A~single temperature offset of 5\,\unit{\degree C} is used. This value was chosen to allow for comparison with previous results (Fig.~\ref{fig:cool05}). In each case, temperature lapse rates are computed using the reference topography from the corresponding dataset.
